---
title: "Data Viz"
---

# **Data Visualization Project**

## **Objective**

This project explores data from two datasets provided by **TidyTuesday**:

1.  **Chess Game Dataset (Lichess):** Analyzing the relationship between chess game duration (number of turns) and player ratings.

2.  **Super Bowl Commercials Dataset:** Examining whether advertisements featuring celebrities receive more views compared to those without celebrities.

## **Dataset Sources**

1.  **Chess Game Dataset** ([Link](https://github.com/rfordatascience/tidytuesday/tree/master/data/2024/2024-10-01))\
    Source: Lichess via TidyTuesday (October 1, 2024)

2.  **Super Bowl Commercials Dataset** ([Link](https://github.com/rfordatascience/tidytuesday/tree/master/data/2021/2021-03-02))\
    Source: TidyTuesday (March 2, 2021)

## TidyTuesday Analysis 1: Chess Game Dataset (Lichess)

### **Objective**

To analyze how the number of turns in chess games varies based on the players' average ratings.

#### Code for Data Processing and Visualization

This section will load the data and create a scatter plot that compares the **number of turns** to **player ranking**. It will help explore how the duration of chess games varies based on the players' ratings.

```{r}
# Load necessary libraries
suppressWarnings({
  library(tidyverse)
})
library(ggplot2)

# Load the chess dataset using tidytuesdayR and suppress messages
suppressMessages({
  tuesdata <- tidytuesdayR::tt_load('2024-10-01')
})
chess <- tuesdata$chess

# Data Cleaning
chess_clean <- chess %>%
  mutate(average_rating = (white_rating + black_rating) / 2)

# Visualization: Scatter Plot of Game Turns vs Average Rating
ggplot(chess_clean, aes(x = average_rating, y = turns)) +
  geom_point(alpha = 0.5, color = "blue") +
  labs(
    title = "Scatter Plot of Chess Game Turns vs Average Player Rating",
    x = "Average Player Rating",
    y = "Number of Turns",
    caption = "Source: Lichess dataset via TidyTuesday, 2024-10-01"
  ) +
  theme_minimal()
```

**Alt text:** A scatter plot showing the relationship between the number of chess game turns and the average player rating. Each point represents a chess game, with the x-axis showing the average rating of the players and the y-axis showing the number of turns. Points are semi-transparent to reduce overlap, with a general spread of game durations across different player ratings.

### 

## TidyTuesday Analysis 2: Super Bowl Commercials Dataset

### **Objective**

To analyze the relationship between Super Bowl commercial view counts and the presence of celebrities in the ads.

#### Code for Data Processing and Visualization

This visualization will focus on analyzing **the relationship between Super Bowl commercial view counts and the presence of celebrities in the ad**. This code generate a boxplot that shows whether ads featuring celebrities get more views.

```{r}
# Load necessary libraries
library(tidyverse)
library(ggplot2)

# Load the Super Bowl dataset and suppress messages
suppressMessages({
  tuesdata <- tidytuesdayR::tt_load('2021-03-02')
})
youtube <- tuesdata$youtube

# Data Cleaning
youtube_clean <- youtube %>%
  filter(!is.na(view_count)) %>%
  mutate(celebrity_presence = ifelse(celebrity == TRUE, "Celebrity", "No Celebrity"))

# Visualization: Boxplot of View Count based on Celebrity Presence
ggplot(youtube_clean, aes(x = celebrity_presence, y = view_count, fill = celebrity_presence)) +
  geom_boxplot() +
  scale_y_log10() +  # Log scale for better visualization
  labs(
    title = "Super Bowl Ad View Count by Celebrity Presence",
    x = "Celebrity Presence in Ad",
    y = "View Count (Log Scale)",
    caption = "Source: Superbowl Commercials dataset via TidyTuesday, 2021-03-02"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

```

**Alt text:** A boxplot comparing Super Bowl commercial view counts based on celebrity presence. The x-axis categorizes ads as 'Celebrity' or 'No Celebrity,' and the y-axis represents view counts on a logarithmic scale. Ads featuring celebrities have a higher median view count compared to those without, with noticeable variability within both categories.

### **Conclusion**

1.  **Chess Dataset:** The scatter plot highlights how the number of turns in chess games correlates with player ratings. Games with highly-rated players tend to have different dynamics compared to lower-rated players.

2.  **Super Bowl Commercials:** The boxplot reveals that ads with celebrities tend to garner more views, showcasing the influence of star power.
