---
title: "Mini Project 2"
---

# Text Analysis of NY Times Headlines

### Data Resource: **RTextTools** Package in R

## Step 1: Load the Dataset & Converting it to a Tibble for Easier Manipulation

```{r loading-dataset}
# Load the package
library(RTextTools)

# Load the NYTimes dataset
data(NYTimes)

# Load the tibble package
library(tibble)

# Convert the NYTimes dataset to a tibble
nytimes_data <- as_tibble(NYTimes)

# Check the structure of the dataset
str(nytimes_data)
```

## Step 2: Exploring the Dataset

```{r exploring-dataset}
# View the first few rows of the dataset
head(nytimes_data)

# Check the structure of the dataset
str(nytimes_data)

# Get a summary of the dataset
summary(nytimes_data)
```

## Step 3: Analysis Using String Functions

```{r}
library(stringr)

# 1. Count the number of headlines that contain the word "politics"
politics_count <- sum(str_detect(nytimes_data$Title, "politics"), na.rm = TRUE)
cat("Number of headlines containing 'politics':", politics_count)

# 2. Remove punctuation from headlines
nytimes_data$clean_headlines <- str_replace_all(nytimes_data$Title, "[[:punct:]]", "")
head(nytimes_data$clean_headlines)

# 3. Extract the first word of each headline
nytimes_data$first_word <- str_extract(nytimes_data$Title, "\\b\\w+\\b")
head(nytimes_data$first_word)
```

## Step 4: Regular Expressions

```{r}
library(stringr)

# 1. Count the number of headlines that contain numbers
nytimes_data$has_numbers <- str_detect(nytimes_data$Title, "\\d+")
num_with_numbers <- sum(nytimes_data$has_numbers, na.rm = TRUE)
cat("Number of headlines containing numbers:", num_with_numbers, "\n")

# 2. Check for broader keywords like "Race" or "Budget"
nytimes_data$contains_race <- str_detect(nytimes_data$Title, "Race")
nytimes_data$contains_budget <- str_detect(nytimes_data$Title, "Budget")

# Summarize counts
num_with_race <- sum(nytimes_data$contains_race, na.rm = TRUE)
num_with_budget <- sum(nytimes_data$contains_budget, na.rm = TRUE)
cat("Number of headlines containing 'Race':", num_with_race, "\n")
cat("Number of headlines containing 'Budget':", num_with_budget, "\n")

# 3. Search for any mention of years from 1990 to 2020
nytimes_data$contains_year <- str_detect(nytimes_data$Title, "19[0-9]{2}|20[0-2][0-9]")
num_with_year <- sum(nytimes_data$contains_year, na.rm = TRUE)
cat("Number of headlines containing years from 1990-2020:", num_with_year, "\n")
```

### Step 5: Visualization

**(a) Bar Plot of Word Frequency**:

```{r}
library(ggplot2)

# Create a data frame for the keyword counts
keyword_data <- data.frame(
  Keyword = c("Numbers", "Race", "Budget", "Year (1990-2020)"),
  Count = c(num_with_numbers, num_with_race, num_with_budget, num_with_year)
)

# Create a bar plot
ggplot(keyword_data, aes(x = Keyword, y = Count, fill = Keyword)) +
  geom_bar(stat = "identity") +
  labs(title = "Keyword Frequency in NY Times Headlines",
       x = "Keyword", y = "Number of Headlines") +
  theme_minimal() +
  theme(legend.position = "none")
```

**Plot Description: Plot (a) highlights the prominence of numbers and certain keywords in the headlines.**

**(b) Pie Chart of Headlines Containing Numbers:**

```{r}
# Create a data frame for the pie chart
number_data <- data.frame(
  Category = c("Contains Numbers", "Doesn't Contain Numbers"),
  Count = c(num_with_numbers, nrow(nytimes_data) - num_with_numbers)
)

# Create the pie chart
ggplot(number_data, aes(x = "", y = Count, fill = Category)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y") +
  labs(title = "Proportion of Headlines Containing Numbers") +
  theme_void() +
  theme(legend.title = element_blank())

```

**Plot Description: Plot (b) highlights the prominence of numbers and certain keywords in the headlines.**

**(c) Line Plot of Year Mentions (1990-2005)**

```{r}
# Extract the mentioned years from headlines (1990-2005
nytimes_data$mentioned_year <- str_extract(nytimes_data$Title, "19[0-9]{2}|20[0-2][0-9]")

# Convert the mentioned_year column to a factor for plotting
nytimes_data$mentioned_year <- factor(nytimes_data$mentioned_year, levels = as.character(1990:2005))

# Filter out any missing or irrelevant years outside the range (1990-2005)
filtered_year_data <- nytimes_data[!is.na(nytimes_data$mentioned_year), ]

# Count the number of mentions of each year
year_count <- table(filtered_year_data$mentioned_year)

# Convert the table to a data frame for plotting
year_count_df <- as.data.frame(year_count)
colnames(year_count_df) <- c("Year", "Count")

# Create a line plot showing year mentions over time (1990-2005)
ggplot(year_count_df, aes(x = Year, y = Count)) +
  geom_line(group = 1, color = "blue", size = 1) +
  geom_point(color = "red", size = 2) +
  labs(title = "Mentions of Years in NY Times Headlines (1990-2005)",
       x = "Year", y = "Number of Mentions") +
  theme_minimal()

```

**Plot Description: Plot (c) provides a historical view of how frequently different years were mentioned in headlines.**
