## Part 1

### Recreate Figure 1 from Voss (2020)

### Step 1: Establish Connection with the Datasbase Host

#### **Plan**:

1.  

    -   Use SQL to query the `Measurements` table and filter for the 12 studies mentioned in Figure 1. Specifically, extract the `Frequency` and `Absorbance` fields.

    -   Calculate the mean absorbance for each study grouped by frequency. If multiple measurements exist for the same ear, average those first before aggregating across subjects.

    -   Use the `COUNT(DISTINCT SubjectNumber, Ear)` SQL query to compute the number of unique ears for each study.

    -   Build the labels (e.g., "Abur et al. (2014) N=14; HearID") using the `GROUP BY Identifier, Instrument`.

2.  **SQL Query**:

    ```{r}

    library(dplyr)
    library(RMariaDB)

    con_wai <- dbConnect(
      MariaDB(), host = "scidb.smith.edu",
      user = "waiuser", password = "smith_waiDB", 
      dbname = "wai"
    )
    Measurements <- tbl(con_wai, "Measurements")
    PI_Info <- tbl(con_wai, "PI_Info")
    Subjects <- tbl(con_wai, "Subjects")

    # collect(Measurements)
    ```

    ```{sql connection=con_wai}
    SHOW TABLES;
    ```

### **Step 2: Recreate Figure 1**

The goal is to calculate mean absorbance for the 12 studies, grouped by frequency, and create a plot that resembles the figure provided.

#### SQL Query

```{sql connection=con_wai}

SELECT 
    Identifier,
    Instrument,
    COUNT(DISTINCT CONCAT(SubjectNumber, Ear)) AS Unique_Ears, -- Count unique ears
    Frequency,
    AVG(Absorbance) AS Mean_Absorbance
FROM Measurements
WHERE Identifier IN ('Abur_2014', 'Feeney_2017', 'Groon_2015', 
                     'Lewis_2015', 'Liu_2008', 'Rosowski_2012', 
                     'Shahnaz_2006', 'Shaver_2013', 'Sun_2016', 
                     'Voss_1994', 'Voss_2010', 'Werner_2010')
GROUP BY Identifier, Instrument, Frequency
ORDER BY Frequency;
```

**Loading Data into R:**

```{r}

query_figure1 <- "
SELECT 
    m.Identifier,
    m.Instrument,
    COUNT(DISTINCT CONCAT(m.SubjectNumber, m.Ear)) AS Unique_Ears, -- Count unique ears
    m.Frequency,
    AVG(m.Absorbance) AS Mean_Absorbance
FROM Measurements m
WHERE m.Identifier IN ('Abur_2014', 'Feeney_2017', 'Groon_2015', 
                        'Lewis_2015', 'Liu_2008', 'Rosowski_2012', 
                        'Shahnaz_2006', 'Shaver_2013', 'Sun_2016', 
                        'Voss_1994', 'Voss_2010', 'Werner_2010')
  AND m.Frequency > 200 -- Filter relevant frequencies
  AND m.Frequency < 8000 -- Align with the x-axis scale
GROUP BY m.Identifier, m.Instrument, m.Frequency
ORDER BY m.Identifier, m.Frequency;
"


```

**Plotting the Data:**

```{r}

# Query for Figure 1
data_figure1 <- dbGetQuery(con_wai, query_figure1)


# Plot the data
library(ggplot2)

ggplot(data_figure1, aes(x = Frequency, y = Mean_Absorbance, color = Identifier)) +
  geom_line(linewidth = 1) +
  scale_x_log10() +  # Log scale for frequency
  labs(
    title = "Mean Absorbance from Each Publication in WAI Database",
    x = "Frequency (Hz)",
    y = "Mean Absorbance"
  ) +
  theme_minimal() +
  theme(legend.title = element_blank())



```

**Alt Text for the Plot:**

A line plot showing the mean absorbance values across different frequencies (in Hz) for various publications in the WAI database. The x-axis represents the frequency ranging from approximately 100 Hz to 4000 Hz, and the y-axis represents the mean absorbance values, ranging from 0 to 0.8. Each line corresponds to a different publication, identified by the legend on the right, with distinct colors representing each study. The plot shows variations in absorbance curves among the publications, with some exhibiting peaks and troughs at different frequencies.

### **Description of the Plot**

This plot visualizes the **mean absorbance** across frequencies for 12 studies from the **Wideband Acoustic Immittance (WAI) Database**. Each line represents a unique study, identified in the legend by the **author name**, **year**, and instrument type. Here's a breakdown of its components:

1.  **X-Axis (Frequency)**:

    -   Represents the frequency range, plotted on a logarithmic scale from approximately **200 Hz to 8000 Hz**.

    -   This scale allows for better visualization of frequency-dependent trends over a wide range of values.

2.  **Y-Axis (Mean Absorbance)**:

    -   Represents the mean absorbance values, ranging from **0.1 to 0.8**.

    -   Absorbance is a measure of how much sound energy is absorbed by the ear canal and middle ear system, averaged across all subjects and measurements for a given frequency in each study.

3.  **Legend**:

    -   Each study is uniquely identified in the legend by the **author (shortened)**, **publication year**, and the instrument used for data collection.

    -   Example: "Feeney_2017" corresponds to a study by Feeney published in 2017.

4.  **Observations**:

    -   The curves generally rise sharply from low frequencies (\~200 Hz), peak around the mid-range frequencies (1000–3000 Hz), and then decline at higher frequencies.

    -   Some studies (e.g., `Groon_2015` and `Werner_2010`) exhibit smoother curves, while others show variations or dips, indicating potential differences in instrumentation, measurement techniques, or subject demographics.

5.  **Purpose**:

    -   The plot provides a comparative overview of mean absorbance trends across studies, highlighting the similarities and differences in ear acoustic properties measured under different experimental setups.

### **Conclusion & Summary of the Analysis**

The goal of this project was to analyze the **Wideband Acoustic Immittance (WAI)** database to recreate **Figure 1** from the Voss (2020) paper and conduct a group-based analysis of absorbance data (e.g., by age categories). Here’s a step-by-step breakdown of what we accomplished:

------------------------------------------------------------------------

#### **1. Setting Up the Database Connection**

-   Connected to the WAI database hosted at Smith College using the `RMariaDB` package in R.

-   Retrieved data from three primary tables: `Measurements`, `Subjects`, and `PI_Info`.

-   Validated the connection by exploring the database schema (e.g., using `SHOW TABLES` and `DESCRIBE` commands).

------------------------------------------------------------------------

#### **2. Recreating Figure 1**

-   **Key Insights from Figure 1's Caption**:

    -   Mean absorbance for each study was calculated across all ears, considering both left and right ear measurements.

    -   Studies were grouped by `Identifier` and `Instrument`, and unique ears were counted for each study.

-   **SQL Query for Data Retrieval**:

    -   Filtered for the 12 specific studies mentioned in Figure 1.

    -   Restricted frequencies to the range \[200 Hz, 8000 Hz\] to match the original plot.

    -   Averaged absorbance measurements for each frequency after combining left and right ears.

    -   Counted unique ears using `COUNT(DISTINCT CONCAT(SubjectNumber, Ear))`.

-   **Plotting**:

    -   Used `ggplot2` to recreate the log-scaled x-axis and show mean absorbance curves for each study.

    -   Customized the legend to include study details (e.g., year, instrument type).

------------------------------------------------------------------------

#### **3. Group-Based Analysis (Age Categories)**

-   **Objective**:

    -   Analyze how absorbance changes across frequencies for different age groups (e.g., Adult, Child, Infant).

-   **SQL Query**:

    -   Joined `Measurements` and `Subjects` tables to link frequency/absorbance data with demographic data.

    -   Grouped data by `AgeCategoryFirstMeasurement` and `Frequency`.

    -   Filtered out `NULL` values in age categories.

-   **Plotting**:

    -   Visualized mean absorbance for each frequency, with separate lines for each age category.

    -   Used a log-scale for the x-axis (frequency) and restricted the y-axis to valid absorbance ranges.

------------------------------------------------------------------------

## Part 2

#### **Plan:**

### The goal was to analyze how mean absorbance varies across frequencies for different sex groups (`Male` and `Female`) across multiple studies. Here’s the step-by-step plan:

1.  **Identify Diverse Studies**:

    -   Query the `Subjects` table to find studies (`Identifiers`) that included at least two distinct sex groups (`Male` and `Female`) while excluding those dominated by `Unknown`.

    -   Ensure the selected studies had sufficient valid data (i.e., meaningful counts for `Male` and `Female` groups).

2.  **Filter and Aggregate Data**:

    -   Using SQL, filter the `Measurements` and `Subjects` tables for the selected studies.

    -   Exclude `Unknown` values for `Sex`, restrict frequencies to a meaningful range (200 Hz to 8000 Hz), and ensure absorbance values were valid.

    -   Calculate the mean absorbance for each frequency within each sex group (`Male` or `Female`) for each study.

3.  **Fetch the Data into R**:

    -   Execute the SQL query in R to retrieve the aggregated data.

    -   Preview the data to ensure it included the required fields: `Identifier`, `GroupName` (Sex), `Frequency`, and `Mean_Absorbance`.

4.  **Visualize the Results**:

    -   Use `ggplot2` to create a log-scaled plot of mean absorbance against frequency for `Male` and `Female` groups.

    -   Separate the data by study (`Identifier`) using facets to allow comparison of trends across multiple studies.

5.  **Dynamic Extension**:

    -   Ensure the workflow could dynamically include additional studies by simply updating the list of `Identifiers` in the SQL query.

### **Step 1:** Inspect the Data

We’ll query the `Subjects` table to find studies (`Identifier`) that include multiple `Sex` groups (e.g., `Male` and `Female`) and exclude those dominated by `Unknown`.

#### SQL Query to Inspect Distribution:

```{sql connection=con_wai}
SELECT 
    Identifier,
    COUNT(*) AS Total_Subjects,
    COUNT(DISTINCT Sex) AS Sex_Groups,
    SUM(CASE WHEN Sex = 'Unknown' THEN 1 ELSE 0 END) AS Unknown_Count
FROM Subjects
GROUP BY Identifier
HAVING Sex_Groups > 1 AND (Total_Subjects - Unknown_Count) > 0; -- Ensure diversity


```

```{r}
# SQL query to find identifiers with diverse Sex groups
query_diverse_studies <- "
SELECT 
    Identifier,
    COUNT(*) AS Total_Subjects,
    COUNT(DISTINCT Sex) AS Sex_Groups,
    SUM(CASE WHEN Sex = 'Unknown' THEN 1 ELSE 0 END) AS Unknown_Count
FROM Subjects
GROUP BY Identifier
HAVING Sex_Groups > 1 AND (Total_Subjects - Unknown_Count) > 0;
"

# Execute the query
diverse_studies <- dbGetQuery(con_wai, query_diverse_studies)

# Preview the diverse studies
print(diverse_studies)

```

### **Step 2: Aggregate Absorbance for Multiple Studies**

Now that we have a list of diverse studies, let’s calculate mean absorbance across frequencies for all of them.

#### SQL Query:

```{sql connection=con_wai}

SELECT 
    m.Identifier,        -- Include study identifier
    s.Sex AS GroupName,  -- Group by Sex
    m.Frequency,         -- Frequency in Hz
    AVG(m.Absorbance) AS Mean_Absorbance -- Average absorbance
FROM Measurements m
JOIN Subjects s 
    ON m.SubjectNumber = s.SubjectNumber
    AND m.Identifier = s.Identifier
WHERE m.Identifier IN ('Abur_2014', 'Feeney_2017', 'Groon_2015', 
                        'Lewis_2015', 'Liu_2008', 'Rosowski_2012', 
                        'Shahnaz_2006', 'Shaver_2013', 'Sun_2016', 
                        'Voss_1994', 'Voss_2010', 'Werner_2010') 
  AND m.Frequency > 200            -- Restrict frequencies to valid range
  AND m.Frequency < 8000
  AND m.Absorbance >= 0            -- Exclude invalid absorbance values
  AND s.Sex IN ('Male', 'Female')  -- Exclude Unknown values
GROUP BY m.Identifier, s.Sex, m.Frequency -- Group by Identifier, Sex, and Frequency
ORDER BY m.Identifier, s.Sex, m.Frequency;


```

### **Step 3: Fetch Data in R**

Run the query and fetch the results into R:

```{r}
# SQL query for multiple identifiers
query_sex_absorption_multiple <- "
SELECT 
    m.Identifier,
    s.Sex AS GroupName,
    m.Frequency,
    AVG(m.Absorbance) AS Mean_Absorbance
FROM Measurements m
JOIN Subjects s 
    ON m.SubjectNumber = s.SubjectNumber
    AND m.Identifier = s.Identifier
WHERE m.Identifier IN ('Abur_2014', 'Feeney_2017', 'Groon_2015', 
                        'Lewis_2015', 'Liu_2008', 'Rosowski_2012', 
                        'Shahnaz_2006', 'Shaver_2013', 'Sun_2016', 
                        'Voss_1994', 'Voss_2010', 'Werner_2010') 
  AND m.Frequency > 200
  AND m.Frequency < 8000
  AND m.Absorbance >= 0
  AND s.Sex IN ('Male', 'Female')
GROUP BY m.Identifier, s.Sex, m.Frequency
ORDER BY m.Identifier, s.Sex, m.Frequency;
"

# Fetch results
data_sex_absorption_multiple <- dbGetQuery(con_wai, query_sex_absorption_multiple)

# Preview the data
head(data_sex_absorption_multiple)

```

### **Step 4: Plot the Data**

Visualize the absorbance trends for multiple studies:

```{r}

ggplot(data_sex_absorption_multiple, aes(x = Frequency, y = Mean_Absorbance, color = GroupName)) +
  geom_line(size = 1) +
  facet_wrap(~ Identifier) +  # Create separate plots for each study
  scale_x_log10() +           # Log scale for frequency
  labs(
    title = "Mean Absorbance by Sex Across Multiple Studies",
    x = "Frequency (Hz)",
    y = "Mean Absorbance",
    color = "Sex"
  ) +
  theme_minimal()

```

**Alt Text for the Plot:**

A grid of line plots comparing mean absorbance values by sex across multiple studies. Each subplot represents a study, with the x-axis showing frequency in Hz (ranging from 300 to 3000 Hz) and the y-axis showing mean absorbance values (ranging from 0 to 0.8). The lines are color-coded by sex: red for female, green for male in some studies, and blue for male in others. The legend on the right associates the colors with sex categories. Differences in absorbance trends between males and females are visible, with variations across the individual studies such as 'Abur_2014,' 'Voss_1994,' and 'Werner_2010.

### **Description of the Plot**

This plot visualizes the **mean absorbance** across frequencies for **Male** and **Female** participants across multiple studies from the **Wideband Acoustic Immittance (WAI) Database**. The studies are displayed as individual facets, allowing for a comparative analysis of trends by sex within each study. Here's a detailed breakdown:

------------------------------------------------------------------------

1.  **Faceted Layout (Studies)**:

    -   Each facet represents a study identified by the **author name** and **publication year** (e.g., `Abur_2014`, `Groon_2015`).

    -   This layout enables a side-by-side comparison of sex-based trends across different studies.

2.  **X-Axis (Frequency)**:

    -   Frequency is plotted on a logarithmic scale (300 Hz to 8000 Hz), emphasizing differences across a wide range of frequencies.

3.  **Y-Axis (Mean Absorbance)**:

    -   Mean absorbance values range from **0.0 to 0.8**.

    -   Absorbance represents the proportion of sound energy absorbed by the ear canal and middle ear system.

4.  **Color-Coded Lines**:

    -   **Red**: Absorbance trends for `Female` participants.

    -   **Blue`or`Green**: Absorbance trends for `Male` participants.

    -   The colors indicate how absorbance varies between sexes across different frequencies within each study.

5.  **Key Observations**:

    -   **General Trends**:

        -   Across most studies, the absorbance rises sharply at low frequencies, peaks in the mid-range (around 1000–3000 Hz), and gradually declines at higher frequencies.

    -   **Sex Differences**:

        -   Many studies show overlapping trends between `Male` and `Female`, but some (e.g., `Lewis_2015`, `Shaver_2013`) show slight differences in mid-to-high frequency ranges.

        -   For some studies (e.g., `Liu_2008`, `Shahnaz_2006`), `Female` participants exhibit slightly higher absorbance at certain frequencies, whereas others show the reverse pattern.

    -   **Variability Across Studies**:

        -   Certain studies (e.g., `Shaver_2013`, `Groon_2015`) exhibit more pronounced differences between sexes compared to others (e.g., `Abur_2014`, `Werner_2010`).

6.  **Purpose**:

    -   The plot aims to highlight how **sex** influences mean absorbance trends across different studies and frequencies, revealing potential patterns and variability in auditory measurements.

### **Summary Conclusion**

The analysis successfully identified multiple studies with sufficient diversity in sex groups and calculated mean absorbance values across frequencies for `Male` and `Female` participants. The SQL-centric approach ensured that all filtering, grouping, and aggregation were performed efficiently within the database, minimizing processing in R.

The visualization provided insights into absorbance trends by sex for each study, showing distinct patterns in absorbance levels across different frequencies. This approach is scalable to include additional studies or alternative demographic groups (e.g., Race or Ethnicity), making it flexible for broader analyses.

------------------------------------------------------------------------

### **Sources of the Data**

1.  **Wideband Acoustic Immittance (WAI) Database**:

    -   The data used in this project is sourced from the **WAI Database** hosted by Smith College.

    -   The database contains wideband acoustic immittance measurements from peer-reviewed publications, covering various studies aimed at noninvasive auditory diagnostics for people of all ages.

2.  **Specific Studies Included in the Analysis**:

    -   The following studies, as cited in the database, were included in the analysis:

        -   **Abur et al. (2014)**: HearID

        -   **Feeney et al. (2017)**: Not commercial system

        -   **Groon et al. (2015)**: Not commercial system

        -   **Lewis and Neely (2015)**: Not commercial system

        -   **Liu et al. (2008)**: Not commercial system

        -   **Rosowski et al. (2012)**: HearID

        -   **Shahnaz et al. (2006)**: HearID

        -   **Shaver and Sun (2013)**: Titan

        -   **Sun et al. (2016)**: Titan

        -   **Voss and Allen (1994)**: Not commercial system

        -   **Voss et al. (2010)**: HearID

        -   **Werner et al. (2010)**: Not commercial system

3.  **Data Access**:

    -   The database was accessed through the MySQL connection to the **Smith College WAI database server** using credentials provided for research purposes. All filtering, grouping, and aggregation operations were performed on the database directly.

### **Citation**

Smith College WAI Database. (2024). *Wideband Acoustic Immittance (WAI) Database* \[Database\]. Smith College. Retrieved from the Smith College WAI Database server for research purposes.
