---
title: "Project 5 Overview: Website Redesign and Impact"
author: "Omar Mnfy"
subtitle: "December 9"
format:
  revealjs:
    scrollable: true
    slide-number: true
    show-slide-number: all
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

## Introduction

-   **Objective:** Showcase the improvements made to the website as part of Project 5.
-   **Focus Areas:**
    1.  Cleaning up and organizing content.
    2.  Enhancing data visualizations and narratives.
    3.  Ensuring proper citations and reproducibility.

------------------------------------------------------------------------

## Before the Redesign

-   **Challenges Identified:**
    -   Warnings and unnecessary messages cluttering the website.
    -   Missing context for analyses and datasets.
    -   Inconsistent or unclear visualization labels.
    -   Lack of proper citations and attributions.

------------------------------------------------------------------------

## After the Redesign

-   **Improvements Made:**
    -   Removed all warnings and messages.
    -   Added context and narratives for each project.
    -   Updated all visualizations with:
        -   Proper titles and captions.
        -   Alt text for accessibility.
    -   Enhanced citations with detailed sources and purpose of datasets.

------------------------------------------------------------------------

## Data Ethics and Documentation

### **Why Ethics Matter in Data Science:**

-   Ensures that data is used responsibly.
-   Protects the integrity of the analysis.
-   Builds trust with stakeholders and audiences.

### **Key Improvements Made:**

1.  **Transparent Methodology:**
    -   Documented all steps taken during the analysis to ensure clarity.
2.  **Privacy Considerations:**
    -   Verified that no personal or sensitive information is included in datasets.

------------------------------------------------------------------------

## Good vs. Bad Documentation Examples

### **Bad Example:**

-   Dataset referenced vaguely with no attribution.

-   Visualization caption: "Ratings distribution."

-   Code comments missing or unhelpful:

    ```{r}
    # Filter data.
    ```

-   No explanation of methods used for analysis.

------------------------------------------------------------------------

### **Good Example:**

-   Detailed dataset citation:

    -   "Source: TidyTuesday Lichess Dataset. Collected by R4DS community from public game archives on Lichess.org for open data exploration."

-   Visualization caption:

    -   "Figure 1: Distribution of player ratings for different time controls (Bullet, Blitz, Rapid) in online chess games. Data source: Lichess."

-   Code comments:

    ```{r}
    # Filter data to include only games played in 2024 in dataset chess_data.
    ```
